---
title: "Final_Project"
author: "Jenny Cribbs"
date: "2026-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Description 

During the summer of 2023, my UC Davis field crew and my collaborator's UCSB field crew surveyed plots randomly selected plots within sugar pine habitat in Yosemite to determine the incidence of white pine blister rust (WPBR). To test whether fire impacted WPBR or sugar pine recruitment, 23 plots were established in the Rim Fire footprint. 

## Data Citation and Availability

National Park Service Inventory and Monitoring Division. 2025. Blister Rust in Sugar Pine Inventory at Yosemite National Park 2023 - Open Format Dataset. National Park Service. Fort Collins CO
https://doi.org/10.57830/2308454

The data is publically available via the NPS IRMA datastore:
<https://irma.nps.gov/DataStore/Reference/Profile/2308454>

## Data Exploration

```{r}
# load packages
library(tidyverse)
library(randomForest)

# load data
yose_plots <- read_csv("data/IRMA_YOSE/YOSE_SugarPine_Event.csv")
# had to re-run the cleaning scripts to figure out an error in the online tree data
yose_trees <- read_csv("/Users/jennifercribbs/Documents/R-Projects/MultipleDisturbances/outputSandbox/YOSE_GBIFoccurrence.csv")
# look at the variables
colnames(yose_plots)
colnames(yose_trees)
```

Since this dataset needed to meet NPS standards, it contains several columns that aren't relevant to this particular study. As explained in the data description, there were two slightly different sampling protocols for this project. The randomly selected plots within sugar pine habitat were evaluated with the primary goal of assessing each tree for WPBR. If there were no sugar pines at the randomly selected point, the plot was shifted up to 500m (this cutoff was to attempt to keep the plot in the same climate strata) or rejected. The high severity fire plots were not shifted since the goal was to understand post-fire effects and regeneraton. One randomly selected plot 48 had no sugar pines, but we recorded data for the trees and understory vegetation since we spent multiple days getting there. Later we found some sugar pines near the trail in the vicinity of point 48. Unfortunately, the trees were so far from the official plot 48 that this really doesn't meet critiera to be considered a plot. Consequently, I am removing it from this dataset. 

```{r}
# data cleaning
# remove irrelevant event-level columns
yose_plots <- yose_plots %>% select(eventID, samplingProtocol, sampleSizeValue_total, startDateTime, verbatimLatitude, verbatimLongitude, verbatimElevation, eventRemarks, slope, aspect)
# remove irrelevant occurrence-level columns
yose_trees <- yose_trees %>% select(-basisOfRecord, -scientificName, -individualCount, -organismQuantity, -organismQuantityType, -publicDisplay, -dataAccess, -lifeStage, -sex, -behavior, -covariateSample, -preparations, -identifiedBy, -identificationReferences, -identificationRemarks, -identificationQualifier, -identificationVerificationStatus, -materialSampleID, -recordNumber, -organismRemarks, -identificationID)

# join to add plot info to the tree table 
yose <- left_join(yose_trees, yose_plots, by = "eventID")

# remove "plot" 48 since it doesn't meet plot criteria
# check that these 34 trees are from plot 48
yose48 <- yose[yose$eventID == 48, ]
# subset to keep only trees in normal plots
yose <- yose[yose_trees$eventID != 48, ]

```


## Data Exploration


```{r}
# overview of the data
summary(yose)
# distributions of the continuous variables
hist(yose$diameter, breaks = 50)
hist(yose$height, breaks = 50)
hist(yose$percentLive, breaks = 50)
```

None of these distributions are normal. Percent live is almost bimodal with quite a few zeros for dead trees and many very healthy trees in the high 90s to 100 percent live. There are trees in between, but it took adding a lot of breaks to even see that. I suspect the trees with WPBR will make up a large part of the relatively rare cases in the middle of the percent live spectrum.

Prior to re-cleaning this data, I noticed a data entry error (a height of 145.3, which is completely out of range). I re-ran the cleaning scripts and fixed that issue. The tooTall object was just a quick check of our extremely tall trees. They may represent some of our least reliable hypsometer measurements due to the difficulty of measuring very tall trees in a crowded forest, but they are not data entry errors. 

```{r}
# look for trees that may be too tall
tooTall <- filter(yose, height >= 80, ) 
# 145.3 was entered for DBH and height--fixed, now just 3 very tall trees
# 3 trees with no height or diameter are actually absences
filter(yose, is.na(height) & is.na(diameter))
# remove absences
yose <- filter(yose, occurrenceID != "E29-PILANA") 
yose <- filter(yose, occurrenceID != "E7-PILANA")
yose <- filter(yose, occurrenceID != "E8-PILANA")
# other missing measurements
filter(yose, is.na(height) | is.na(diameter))
```

```{r}
# check correlations between a few key predictors
cor(yose$diameter, yose$height, use = "complete.obs")
cor(yose$diameter, yose$percentLive, use = "complete.obs")
cor(yose$height, yose$percentLive, use = "complete.obs")

```

As expected there is a strong correlation between tree girth and height. There is a weak negative correlation between tree girth and percent live, and not much of a correlation between height and percent live. Very large diameter trees tend to be older, and may simply have had more time to get attacked by pests and pathogens or get damaged mechanically. 

```{r}
# visualize the relationship between height and diameter and species
species <- ggplot(yose, aes(diameter, height, color = taxonID)) +
  geom_point(alpha = .2)
ggsave("species.png", species)
```
There don't seem to be obvious species effects other than the fact that sugar pines represent the widest and tallest trees. This makes sense biologically, but also sugar pine was the target species, meaning we measured all mature individuals in sometimes large variable area plots in an effort to get at least 20 sugar pines. The other species were only measured in a smaller 50 by 20m plot. The oaks also don't get as tall as the pines, and resprout after fire. Their very different growth habit will likely not improve prediction for missing conifer values, so I am filtering out the three species of oaks and California bay laurel in order to explore allometric relationships for conifers. 

```{r}
# get a list of unique tree species
unique(yose$taxonID)
# remove non-conifer species
yose_cons <- yose %>% filter(taxonID != "UMCA") %>% 
  filter(taxonID != "QUKE") %>% 
  filter(taxonID != "QUCH") %>% 
  filter (taxonID != "QUWI")
```


Compare imputation methods using 1) a published allometric equation for sugar pines in nearby Oakhurst, CA (Zhao et al 2012) 2) simple linear regression, 3) log-log regression based on the idea of power law relationships for tree growth 4) a non-linear asymptotic function, and 5) an unsupervised PCA approach.  

```{r}
# remove actual missing cases and saplings
complete <- yose %>% 
  dplyr::filter(!is.na(height), !is.na(diameter), diameter > 0, height > 0) 
# set seed
set.seed(42)
# use 80% of the data for training
train_ids <- sample(seq_len(nrow(complete)), 0.8*nrow(complete))
test_ids  <- setdiff(seq_len(nrow(complete)), train_ids)
# create a train/test split
train <- complete[train_ids, ]
test  <- complete[-train_ids, ]

```


```{r}
# model 1: Zhao et al 2012
# define a function based on the zhao regression equation for sugar pine
zhao <- function(diameter) {height <- (diameter - 4.37) / 1.81
return(height)
}

# model 2: simple linear regression 
# simple linear regression
slr <- lm(height ~ diameter, data = train)
# look at coefficients
summary(slr)
# model diagnostic plots 
plot(slr)
# plot height versus diameter with regression line
plot(train$diameter, train$height)
abline(slr, col = "red")

# model 3: log-log model
llr <- lm(log(height) ~ log(diameter), data = train)
# model summary
summary(llr)
# diagostic plots
plot(llr)
plot(log(train$diameter), log(train$height))
abline(llr, col = "blue")


```

```{r}
# model 4: nonlinear asymptotic model
nls_fit <- nls(
  height ~ 1.3 + a * (1 - exp(-b * diameter)),
  data = complete,
  start = list(a = 60, b = 0.02)
)

```

The overall fit looks pretty good, though it might be improved a bit by accounting for the fact that trees grow taller a little faster than they grow wider. The regression line falls a bit below the bulk of the data points from zero to a little less than 100cm in diameter. 

## predict missing heights

```{r}
# predict with Zhao 2012 equation
zhao_height <- zhao(test$diameter)
# predict with slr model
slr_height <- predict(slr, newdata = test)
# predict with llr model
llr_height <- exp(predict(llr, newdata = test))
# predict with nls model
nls_height <- predict(nls_fit, newdata = test)
```

```{r}
# PCA imputation 
library(missMDA)
# select numeric variables for PCA
vars_numeric <- yose %>% dplyr::select(where(is.numeric))
# center, scale and impute
pca_imp <- imputePCA(vars_numeric, scale = TRUE)
# convert PCA output to a dataframe
pca_complete <- as.data.frame(pca_imp$completeObs)
# extract test rows only for comparison
test$pca_height <- pca_complete[test_ids, "height"]
```

## Calculate Prediciton Accuracy 

```{r}
# create a funciton to calculate RMSE
rmse <- function(obs, pred) sqrt(mean((obs - pred)^2))

# compare models
rmse(test$height, zhao_height)
rmse(test$height, slr_height)
rmse(test$height, llr_height)
rmse(test$height, nls_height)
rmse(test$height, test$pca_height)

rmse_table <- data.frame(
  Model = c("Zhao","SLR","LLR","NLS","PCA"),
  RMSE = c(
    rmse(test$height, zhao_height),
    rmse(test$height, slr_height),
    rmse(test$height, llr_height),
    rmse(test$height, nls_height),
    rmse(test$height, test$pca_height)
  )
) %>% arrange(RMSE)

library(flextable)
flextable(rmse_table)
save_as_image(rmse_flextable)

AIC(slr, llr, nls)
                
```
```{r}
# residual plot for zhao model
plot(test$diameter, test$height - zhao_height)
abline(h = 0)
# residual plot for slr
plot(test$diameter, test$height - slr_height)
abline(h = 0)
# residual plot for log-log model
plot(test$diameter, test$height - llr_height)
abline(h = 0)
# residual plot for nls model
plot(test$diameter, test$height - nls_height)
abline(h = 0)
# residual plot for pca
# residual plot for slr
plot(test$diameter, test$height - test$pca_height)
abline(h = 0)
```


```{r}
# Linear regression CI
slr_pred <- predict(slr, newdata = test, interval = "confidence", level = 0.95)
test$slr_fit <- slr_pred[, "fit"]
test$slr_lwr <- slr_pred[, "lwr"]
test$slr_upr <- slr_pred[, "upr"]

# Log-log regression CI (remember to back-transform)
llr_pred <- predict(llr, newdata = test, interval = "confidence", level = 0.95)
test$llr_fit <- exp(llr_pred[, "fit"])
test$llr_lwr <- exp(llr_pred[, "lwr"])
test$llr_upr <- exp(llr_pred[, "upr"])
```


```{r}
library(ggplot2)

modelComp <- ggplot(test, aes(x = diameter, y = height)) +
  geom_point(alpha = 0.2) +
  geom_line(aes(y = zhao_height), color = "yellow") +
  geom_line(aes(y = slr_height), color = "red") +
  geom_ribbon(aes(ymin = slr_lwr, ymax = slr_upr), alpha = 0.2, fill = "red") +
  geom_line(aes(y = llr_fit), color = "blue") +
  geom_ribbon(aes(ymin = llr_lwr, ymax = llr_upr), alpha = 0.2, fill = "blue") +
  geom_line(aes(y = nls_height), color = "green") +
  theme_minimal() +
  labs(y = "Height", x = "Diameter", title = "Comparison of Height Predictions")

ggsave("modelComparison.png", modelComp)
```

I experimented with height imputation using four approaches: a published allometric equation (Zhao et al 2012), linear regression, log–log allometric regression, nonlinear asymptotic allometric regression, and PCA-based imputation. Predictive accuracy was evaluated by masking observed heights and comparing predicted values to true measurements using RMSE. For the three regression methods, I used 80% of the data for training while reserving 20% as the test set. The Zhao equation did not require training data since it was based on prior work, though I used it to predict the heights for the same test set. For PCA-based imputation, I used the full dataset. 



## WPBR Survey

Consider what to do with the almost bimodal percent live variable and beetle and blister rust observations. Presence/absence is the easiest for both, but the raw data is actually a combination of various signs (beetles: pitch tubes and/or exit holes for one of two species WPBR: active/inactive branch/bole canker counts).

```{r}
# random forest requires complete cases

# use only complete cases for now
yose_clean <- yose_clean[complete.cases(yose_clean), ]

# sensitivity analysis--how sensitive is the imputation of 0.5 vs. something else

# split data in training (700 observations) and testing (300 observations) 
i <- sample(960)[1:700]
train <- yose_clean[i, ]
test  <- yose_clean[-i, ]

# fit a random forest model without tuning
rf <- randomForest::randomForest(train$percentLive~., data=train,  importance=TRUE)

```

## ISLR Chapter 10 Exercise 7

Fit a neural network to the Default data. Use a single hidden layer
with 10 units, and dropout regularization. Have a look at Labs 10.9.1–
10.9.2 for guidance. Compare the classification performance of your
model with that of linear logistic regression.

```{r}
## A Single Layer Network with 10 units on the Default data

###
# load textbook data
library(ISLR2)
# remove NAs from the default data
Gefault <- na.omit(Default) # no NAs?
# store the number of rows in n
n <- nrow(Gefault)
# set a random seed for reproducibility
set.seed(13)
# set aside test data
ntest <- trunc(n / 3)
testid <- sample(1:n, ntest)

### fit a logistic model to predict salary
glmfit <- glm(default ~ ., 
              data = Default[-testid, ], 
              family = binomial)

probs <- predict(glmfit, Default[testid, ], type = "response")

#pred_class <- ifelse(probs > 0.5, "Yes", "No")

#mean(pred_class != Default$default[testid])

### set up x and y for glmnet 
# set up a matrix with scaled predictor variables
x <- scale(model.matrix(default ~ . - 1, data = Default))
# salary as response variable
# check levels of factors for y
levels(y)
# code Yes as 1 and no as 0 (glm net requires numeric)
y_num <- ifelse(y == "Yes", 1, 0)

### fit the lasso using glmnet
library(glmnet)

# lasso regularization 
cvfit <- cv.glmnet(x[-testid, ], y_num[-testid], family = binomial, type.measure = "deviance")
cpred <- predict(cvfit, x[testid, ], s = "lambda.min", type = "response")
pred_class <- ifelse(cpred > 0.5, 1, 0)
mean(pred_class != y_num[testid] - cpred)

### set up neural network structure with pipes between layers
library(keras3)
## reticulate::use_condaenv(condaenv = "r-tensorflow")
modnn <- keras_model_sequential(input_shape = ncol(x)) |>
  layer_dense(units = 10, activation = "relu") |> # 10 hidden units
  layer_dropout(rate = 0.4) |> 
  layer_dense(units = 1, activation = "sigmoid")
# add details to control fit and minimize squared error loss
compile(modnn, loss = "binary_crossentropy", optimizer = optimizer_rmsprop(),
        metrics = list("accuracy"))

### fit the model (this one takes a while 1500 epochs!)
history <- fit(modnn, x[-testid, ], y_num[-testid], epochs = 1500, 
               batch_size=32, validation_data=list(x[testid, ], y[testid]))
# load ggplot2
library(ggplot2)
# visualize the mean error for test and train data
plot(history)
# evaluate final model on test data
npred <- predict(modnn, x[testid, ])
pred_class <- ifelse(npred > 0.5, 1, 0)
# mean salary prediction
mean(pred_class != y_num[testid])

```

```{r}
# logistic regression model for comparison
glm.fits <- glm(default ~ student + balance + income, data = Default , family = binomial)
# look at the logistic regression model
summary(glm.fits)
# use logistic regression model to predict for test data
lpred <- predict(glm.fits, Default[testid, ], type = "response")

# calculate AUC for logistic and neural net models
library(pROC)
# calculate AUC
glm_auc <- roc(y_num[testid], lpred)$auc
nn_auc  <- roc(y_num[testid], npred)$auc

# compare AUC 
glm_auc
nn_auc
```

